---
title: 'Progress Report: C++ Core Sampler Implementation for `dirichletprocess`'
author: "Priyanshu Tiwari"
date: "June 8, 2025"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### **1. Executive Summary**

This report details the progress made in the project to re-implement the core samplers of the package in C++. The primary work done was to port the existing R implementations of all distribution-specific samplers to C++ to achieve substantial performance gains.

The C++ porting of all target distributions is now complete. The new backend has been integrated with the existing R S3 classes, and a comprehensive test suite has been executed. The results are good, with the vast majority of tests passing, which validates the correctness of the new C++ logic. While a small number of tests have failed and require debugging, next step is debugging existing errors, performance optimization and finalize documentation.

---

### **2. Implementation Status**

Porting the samplers to C++—has been achieved.

* **Completed C++ Implementations**: The following distributions and models now have C++ backends:
    * Normal (Conjugate)
    * Multivariate Normal (Conjugate)
    * Beta (Conjugate & Non-conjugate)
    * Exponential (Conjugate)
    * Weibull (Non-conjugate)
    * Hierarchical Normal
    * Hierarchical Beta
    * Hidden Markov Models (HMM)

* **R/C++ Interface**: The `Rcpp` and `RcppArmadillo` libraries were used to create a seamless interface between the R front-end and the C++ back-end. Wrapper functions have been created to ensure that the user experience remains consistent and that the C++ code can be called directly from the existing R functions.

---

### **3. Testing and Validation**

A full test suite was run using `devtools::test()` to validate the correctness of the C++ implementations against the established R versions.

The overall results are as follows:

* **Total Tests Run:** 1515
* **Passed** 1482 
* **Failed** 17 
* **Skipped:** 8 

The high pass rate confirms that the C++ logic is fundamentally sound and produces results consistent with the original R code for a wide range of scenarios.

The 25 failing tests are now the highest priority. Failures are concentrated in a few areas, exponential sampler (`test-exponential-cpp.R`) and statistical comparisons in the hierarchical beta model (`test-hierarchical-beta-cpp.R`). Currently working on it.

---

### **4. Performance Benchmarking: R vs. C++**

The following code provides a template for benchmarking the C++ samplers against their R counterparts.

#### **Example: Gaussian Distribution**

We first compare the `Fit` function for the Gaussian Dirichlet Process.

```{r benchmark_code, eval=TRUE}
# Load necessary libraries
library(dirichletprocess)
library(microbenchmark)
library(ggplot2)

# --- Benchmark Example: Normal Distribution ---
# Generate some sample data for the model
set.seed(123)
# Ensure data is explicitly a matrix.
data <- matrix(rnorm(150, 2, 0.5), ncol = 1)

# 1. Prepare the R version by globally disabling C++
options(dirichletprocess.use_cpp = FALSE)
dp_r <- DirichletProcessGaussian(data, g0 = c(0, 1, 1, 1))
dp_r <- Fit(dp_r, 100) # Initial fit

# 2. Prepare the C++ version by globally enabling C++
enable_cpp_samplers()
dp_cpp <- DirichletProcessGaussian(data, g0 = c(0, 1, 1, 1))
dp_cpp <- Fit(dp_cpp, 100) # Initial fit

# 3. Run the benchmark
# We compare running 1000 MCMC iterations for each implementation.
bench_results <- microbenchmark(
  `R Fit`   = Fit(dp_r, 1000),
  `C++ Fit` = Fit(dp_cpp, 1000),
  times = 10
)

# Print the results table
print(bench_results)

# Visualize the benchmark results
autoplot(bench_results) + 
  ggtitle("Performance Comparison: R vs. C++ for Gaussian DP") +
  theme_light()

# Clean up the global option
options(dirichletprocess.use_cpp = TRUE)
```

#### **Example: Beta Distribution**

```{r benchmark_code_beta, eval=TRUE}
# Load necessary libraries
library(dirichletprocess)
library(microbenchmark)
library(ggplot2)

# --- Benchmark Example: Beta Distribution ---
set.seed(101)
# Generate sample data
beta_data <- matrix(rbeta(100, 2, 5), ncol = 1)

# Define the prior parameters for the Beta model (Shape1 and Shape2)
prior <- c(1.0, 1.0)

# 1. Prepare the R version by globally disabling C++
options(dirichletprocess.use_cpp = FALSE)
dp_beta_r <- DirichletProcessBeta(beta_data, g0 = prior, maxY = 1)
dp_beta_r <- Fit(dp_beta_r, 100) # Initial fit

# 2. Prepare the C++ version by globally enabling C++
enable_cpp_samplers()
dp_beta_cpp <- DirichletProcessBeta(beta_data, g0 = prior, maxY = 1)
dp_beta_cpp <- Fit(dp_beta_cpp, 100) # Initial fit

# 3. Run the benchmark
bench_results_beta <- microbenchmark(
  `R Fit (Beta)`   = Fit(dp_beta_r, 1000),
  `C++ Fit (Beta)` = Fit(dp_beta_cpp, 1000),
  times = 10
)

# Print the results table
print(bench_results_beta)

# Visualize the benchmark results
autoplot(bench_results_beta) + 
  ggtitle("Performance Comparison: R vs. C++ for Beta DP") +
  theme_light()

# Clean up the global option
options(dirichletprocess.use_cpp = TRUE)
```

---

### **5. How to Switch Between R and C++**

The package provides a simple mechanism to choose between the original R implementation and the new C++ backend by setting a global option for the R session.

The main helper function is `enable_cpp_samplers()`. To revert to the R implementation, the option must be set manually.

```{r switch_code}
# Load the package
library(dirichletprocess)

# By default, or after calling enable_cpp_samplers(), new objects will use the C++ backend.
enable_cpp_samplers()
dp_object_cpp <- DirichletProcessGaussian(matrix(rnorm(100), ncol = 1))
# The using_cpp() function checks the global session setting, not a specific object.
cat("Is the session set to use C++? ->", using_cpp(), "\n")


# To create an object with the R implementation, disable the global option first.
options(dirichletprocess.use_cpp = FALSE)
dp_object_r <- DirichletProcessGaussian(matrix(rnorm(100), ncol = 1))
cat("Is the session set to use C++? ->", using_cpp(), "\n")


# The function `using_cpp()` can be used to check which implementation 
# an object is configured to use.

# Re-enable for future work
enable_cpp_samplers()
```

This design allows for easy comparison and debugging by specifying the desired backend for the session before creating the models.

---

### **6. Next Steps**

1.  **Debug Failing Tests**: Investigate and resolve the 17 failing tests. 
2.  **Address Skipped Tests**: Review the 8 skipped tests and implement them where appropriate.
3.  **Update Documentation**: Ensure all new C++ functionalities and the switching mechanism are clearly documented for users.
